\mytitle

### 2022-11-16 Article.tex
Revision

\def\mytitle{Whirlwind: Overload protection, fault--tolerance and self--tuning in an Internet services platform}
\def\myauthor{Peter Donald, Samar Singh and Somnath Ghosh}
\def\mykeywords{concurrency, software architecture, overload, fault-tolerance, self-tuning}

\documentclass[conference]{IEEEtran}

\usepackage{cite}
\usepackage[pdftex]{graphicx,hyperref}
\usepackage{url}
\usepackage{epstopdf}

@@

\newcommand{\todo}[1]{{\bf\begin{center}TODO: #1\end{center}}}

@@

\author{\IEEEauthorblockN{\myauthor}
\IEEEauthorblockA{School of Computer Science and Electronic Engineering\\
La Trobe University,\\
Melbourne, Australia\\
Email: peter@realityforge.org, S.Singh@latrobe.edu.au and S.Ghosh@latrobe.edu.au}
}

@@

\begin{abstract}
Performance and availability are critical importance when Internet services are integrated into emergency response management. Poor performance or service failure can result in severe economic, social or environmental cost. This paper presents Whirlwind, a software architecture that includes primitives for overload management and fault tolerance. A Whirlwind service is composed of a collection of isolated, independent, sequential processes that communicate through asynchronous message passing. If a process fails, the fault is contained within the process and a message is propagated to monitoring processes that may attempt to recover from the error. Processes are grouped with other processes that share similar resource, computation and concurrency requirements. Each group contains a scheduler and a thread pool that drives execution of processes within the group. The group may also define a message predicate that determines if a message posted to a process in the group is accepted. A rejected message typically signals overload and allows the application the chance to perform load shedding and avoid overcommitment of resources. Principals are shared between processes in different groups, enabling consistent prioritization and admission control across groups. The resource management policies are typically driven by feedback loops that monitor resource availability and system performance, and adjust tuning parameters to meet performance goals. Whirlwind evolved over a period of five fire seasons as part of emergency response software in Victoria, Australia.
\end{abstract}

@@

Whirlwind was born out of the need to create a reliable, scalable software architecture for the Incident Resource Information System (IRIS). IRIS tracks the dispatch and deployment of resources during an emergency situation. The resources include everything from vehicles to personnel and the emergency situations have included wildfires, floods and virus outbreaks. IRIS is also responsible for generating the payments for personnel during emergency situations.

%When data is updated in an \emph{area}, the data is replicated to clients who have subscribed to that particular \emph{area}. During emergency situations the data churn tends to be restricted to specific areas that correspond to geographic locations that the emergency occurred in and most clients have subscribed to that particular area. This typically results in a load that grows exponentially with respect to the number of concurrent users.

Internet traffic is bursty~\cite{Crovella:1996:BurstyTraffic} and popular Internet services are likely to be subject to periods of transient or sustained overload. This is intensified for Internet services used in emergency responses. Overload can be caused by unexpectedly high request rates or unexpected shifts in the characteristics of the request distribution. In IRIS the request rate increased dramatically during shift change over times or prior to payment cut--off dates. The characteristics of the requests also changed dramatically depending on whether the users were primarily interacting with the payment subsystem or the resource tracking subsystem.

@@

Whirlwind is based on the actor model of concurrency. It incorporates existing work on load management, fault--tolerance, scheduling and self--tuning and synthesizes these elements into a consistent framework. Whirlwind was developed for the Java Virtual Machine, primarily targeting the Java programming language. The architecture needed to take advantage of the increased number of processing units in modern hardware.

Early in the evolution of Whirlwind it was discovered that it was necessary to provide flexibility in certain aspects such as admission control policies and scheduling algorithms. Plugin points are defined where flexibility is required. To maintain acceptable performance the plugins are designed with coarse--grain interfaces with minimal virtual calls along critical paths.

Included with Whirlwind is an library designed to evaluate and report the performance of Whirlwind applications while avoiding many of the common pitfalls of Java performance evaluation~\cite{Blackburn:EvaluationMethodology,Georges:JavaPerformance}. Part of the goal was to compare different versions of Whirlwind but it is also useful in evaluating a Whirlwind application.

Each iteration of Whirlwind has undergone an extensive performance evaluation before being rolled out into production. Space does not permit presenting a a detailed performance report. Suffice to say that Whirlwind, at least with certain plugins configured, met the original performance goals and compares favorably with other popular production frameworks.

@@

Past research has tended to focus one categorization axis rather than categorizing a request along multiple dimensions. It is probable that this is done as it makes the mapping between a categorization and a priority simple; \emph{Premium} users are always prioritized above \emph{Basic} users, \emph{Buy} requests are always prioritized above \emph{Browse} requests, while during overload light--weight requests are always prioritized above heavy--weight requests etc.

@@

%Faults occur, yet most systems are not designed to expect faults and limit the impact of faults when they occur. Fault tolerant systems must be able to isolate faults, detect a fault has occurred and reinitialize a subsystem after it fails from some known good state~\cite{Schneider:FailStopProcessors}. The system can then retry the operation that lead to the failure. Shared memory architectures can not isolate faults thus limiting fault tolerance within the application~\cite{Armstrong03Thesis}.

A fault tolerant system is one ``\ldots able to recover from errors and to restore normal operation \ldots''~\cite{Renzel:03:ErrorHandling} If there is a decrease in service quality, the degradation is proportional to the severity of the fault. Fault tolerance is achieved by expecting faults and designing the system so that it behaves reasonably in the presence of faults.

The Tandem NonStop system has a mean time between failures measured in years~\cite{Grey85Fail} and indirectly influenced many of the design decisions of Whirlwind. The architecture consists of hierarchically decomposing a system into processes where each process is a single unit of service and a single unit of failure. Processes share no state and only communicate via messages. A failure in a process does not propagate to other processes thus resulting in fault containment. The system was also designed using a fail--fast philosophy so that processes either function correctly or detect the fault, signal failure and stop operation.

The construction of a fault--tolerant computing systems is simplified when programs do not need to cope with arbitrary behavior or corrupted state information. So not only should a failure be contained within a process, but on detection of a failure the process should halt rather than make erroneous state visible to other processes. Other processes should be able to detect if a process has failed so that they can attempt error recovery. Error recovery should have a known good state with which to restart the process. Similar properties are deemed necessary to construct fault--tolerant hardware systems~\cite{Schneider:FailStopProcessors}.

@@

It is argued that  the inherent non--determinism make any non--trivial multi--threaded, shared memory application ``incomprehensible to humans''~\cite{Lee06ThreadProblems}. Multi--threaded, shared memory architectures also face performance challenges. Maintaining throughput and fairness in the presence of locks can be difficult and is highly dependent on lock granularity and expected load patterns.

@@

Erlang code is executed by a specialized runtime. However other actor implementations (i.e. SALSA~\cite{Varela:SALSA}, Scala's Actor library~\cite{Haller:07:ScalaActors}, Kilim~\cite{Srinivasan:08:Kilim}) target a general purpose virtual machine such as the Java Virtual Machine (JVM). These implementations use the type system or code modification techniques to verify that the actors conform to certain rules (i.e. no access to shared memory) and to integrate the actors with a runtime library. The runtime library provides execution environment for the actors and maps the actors to threads using the 1--to--1 model~\cite{Varela:SALSA}, M--to--N model~\cite{Srinivasan:08:Kilim,armstrong96erlang} or a combination of both approaches~\cite{Haller:07:ScalaActors}.

@@

%The resource management policies are typically driven by feedback loops that monitor resource availability and system performance and adjust tuning parameters to meet performance goals

@@

Processes are isolated and do not share memory. To communicate with a process a message is posted to the process. The ownership of the message contents is transferred to the receiving process. Message delivery is asynchronous and messages are placed in a process--local queue or inbox. Message transmission is asynchronous to ensure that the sender is isolated from the receiver and the sender can not be blocked, or delayed by a slow receiver. It should be noted that message passing in Whirlwind obeys the \emph{happened--before} ordering of events~\cite{lamport78}.

@@

% A rogue process can still consume an executor

@@

Most Internet services support many concurrent requests and many of these requests perform similar computations. Identifying requests that are at a similar stage of processing and executing a ``cohort'' of them sequentially increases code and data locality. This in turn increases performance by reducing cache misses~\cite{Larus:2002:Cohort,welsh03Adaptive}. Groups are designed to identify processes that have a high degree of code and data locality and thus the group defines ideal ``cohort''.

@@

The post operation will raise an exception when a message is rejected thus allowing the source process the opportunity to attempt alternative action. A HTTP service may first attempt to deliver a media rich page. If this message is rejected, the service attempts to deliver a lower resolution page. This allows the service to deliver progressively lower resolution content as the load on the service increases thus avoiding overload~\cite{abdelzaher99web}. SEDA demonstrated the effectiveness of the message predicate primitive in overload management~\cite{welsh03Adaptive}.

@@

Whirlwind provides several controller implementations that dynamically adjust the concurrency level of a group. Periodically the controller observes performance characteristics and adjusts the pool size in an attempt to meet throughput and response time goals. The developer can also specify a custom controller or use one of the predefined controllers. The pseudo code for the default algorithm is presented in Figure~\ref{figure:epc}.

\begin{figure}[ht]
\begin{verbatim}
if idleTime > idleTimeThreshold
  if threadCount > minThreadCount
    threadCount--
else if threadCount < maxThreadCount
  if queueLatency > queueLatencyThreshold
    if lockLatency < lockLatencyThreshold
      threadCount++
\end{verbatim}
\caption{Pseudocode for default algorithm for adjusting concurrency level of a group. \emph{idleTime} is the average time a thread waits for the scheduler to supply a process. \emph{queueLatency} is the average time a message spends in the queue. \emph{lockLatency} is the average time spent acquiring a process lock prior to message handling.}
\label{figure:epc}
\end{figure}

@@

As in Erlang, the supervisor processes are typically organized into a tree such that all but the root supervisor processes is monitored by another supervisor. If a supervisor fails to recover from a child failure then it can in turn fail and force the parent supervisor process to attempt error recovery. For the system to fail, all supervisors in the tree must fail to recover from the error.

@@

The best way to get a feel for the architecture is to look at the design for a Whirlwind application. This section describes the design of the HTTP server component used in IRIS. Figure~\ref{figure:http-server} presents a basic design of subsystem responsible for processing HTTP connections.

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{http-server.eps}
\caption{The subsystem responsible for processing HTTP connections. All of the groups contain processes of a single type except for the ``Resource Handler'' group that includes processes of type ``List'' and ``Show'' for listing a set of resources and presenting detailed information about a single resource respectively. For the sake of simplicity, the message paths are presented as flowing in one direction but all paths are bidirectional except the handlers that communicate directly with \emph{HttpConnection} processes.}
\label{figure:http-server}
\end{figure*}

Each connection to the HTTP server resulted in the creation of three processes;

\begin{itemize}
	\item a \verb+SocketConnection+ process to manage the underlying socket,
	\item a \verb+HttpConnection+ process to decode and encode HTTP messages and manage the HTTP connection state, and
	\item a \verb+Dispatcher+ that routes the HTTP requests to specific handlers based on the contents of the headers, method and uri of the request.
\end{itemize}

Each request submitted to the server results in the \verb+Dispatcher+ creating a \emph{handler} process. A handler is responsible for delivering content from the filesystem, dynamically generating content based based on values retrieved from a database or submitting data to an external system.

The handler will usually generate one or more processes while handling the request. For example, the \verb+FileHandler+ process creates an \verb+AsyncFile+ process while the \verb+ListResources+ process creates a \verb+DbConnection+ process.

Handlers will share a group if they utilize similar resources and load shedding is likely to handle these handlers similarly. However, most handlers and other other processes in the system will have a single group per process type.

Most Whirlwind applications will only place messages predicates on a few specific groups. The HTTP server defines a message predicate on the \verb+Dispatcher+ group and on each of the handler groups. If a message predicate attached to a handler group rejects a message then the dispatcher may attempt to send a less resource intensive request to the handler, as presented in Figure~\ref{figure:message-predicate}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{message-predicate.eps}
\caption{The message predicate attached to the \emph{FileHandler} group rejects a message requesting high quality version of a file but accepts a message requesting a low quality version of a file that is less resource intensive.}
\label{figure:message-predicate}
\end{figure}

If the message predicate attached to the \verb+Dispatcher+ group rejects a message then the \verb+HttpConnection+ sending the message will return a ``503 Service Unavailable'' response to the client and close the connection. This message predicate is only triggered when the message predicates attached to the handlers have failed to shed enough load to avoid system overload.

@@

The creation of a new handler for every request rather than handler reuse is deliberate. This design results in all the processes in Figure~\ref{figure:http-server} being  associated with a single principal thus allowing consistent treatment by message predicates across the whole system.

The prioritization of principals takes into account whether the session is for an authenticated user and if the user is in the \emph{admin} role. Administrators are given the highest priority as they need to be able to view the status of the system at anytime, even if the system is overloaded while unauthenticated users are given the lowest priority.

@@

Whirlwind has gone through 5 major revisions. The initial version was primarily focused on effective overload management. The following three revisions were attempts to remove sources of bugs that were revealed once the system was put into production. The latest version focused on improving performance.

\subsection{Whirlwind v1.0}

Whirlwind v1.0 was an event--driven architecture that decomposed the application into a network of stages connected with request queues. Each stage was serviced by a pool of threads. Each thread continuously looped, retrieving batches of events from the request queue and passed them to an event handler. Event handlers could be stateless and shared across all threads. The event handlers could also be stateful and instantiated once for each thread or event. The thread pool size and batch size were adaptively tuned to maximize throughput.

Whirlwind v1.0 was heavily influenced by the SEDA~\cite{welsh03Adaptive} but adapted to use modern JVM technologies such as the Java Management eXtensions (JMX) and Java's New I/O (NIO) toolkit. The implementation was complimented with an extensive automated test suite.

\subsection{Whirlwind v2.0}

SEDA did not provide any explicit mechanisms for protecting shared state from other threads accessing it, either within a single stage or across stages. During the first fire season, several faults were revealed that were traced to poorly accessed shared state.

Whirlwind v2.0 attempted to address this concern by encapsulating shared state in \emph{elements}. Some elements allowed multiple concurrent threads while \emph{sequential} elements only allowed one thread access at any one time. The concurrency constraints on elements were enforced by a byte code validator. The design of \emph{elements} meant that stateful event handlers were no longer necessary.

It should be noted that scheduling decisions across groups were no longer isolated. One stage could acquire a lock and block progress of a thread in another stage. Whirlwind v2.0 also removed event batching as we were unable to reproduce any measurable performance benefit.

\subsection{Whirlwind v3.0}

Whirlwind v2.0 did not guarantee any particular event ordering. Multiple threads could be attempting to delegate event handling to a single element at any one time and the ordering is determined by the JVM or operating system. Mid fire season, the hardware platform was upgraded to 16 processors which increased the frequency of out--of--order event processing and exposed several bugs with some elements ability to handle reordered events.

Whirlwind v3.0 addressed this problem by adding a local FIFO queue to \emph{sequential} elements. Each time an element was scheduled it performed event handling for the next event. As non--sequential elements were not widely used they were removed, effectively transforming Whirlwind into an implementation of the actor model of concurrency. This change cascaded through the system allowing the removal of event handlers and enhancing the adaptive algorithm used to manage the thread pool size.

\subsection{Whirlwind v4.0}

During the preceding fire season, a crash occurred as a result of accidental state sharing between elements. To address these concerns two changes were made to Whirlwind; the Erlang model of fault propagation and recovery was incorporated, and the byte code validator was enhanced to enforce isolation properties. The other major change was one of terminology; elements became processes, events became messages and stages became groups.

To enforce isolation properties of the messages, the byte code validator required that developer annotate the messages. The validator used the annotations combined with rules and heuristics to enforce the isolation constraints. The messages and all transitively referenced values needed to be either; immutable, a process or have ownership transferred from the source to destination processes. The byte code validator in many ways resembled a less capable, more ad--hoc version of the Kilim byte code weaver~\cite{Srinivasan:08:Kilim} that just performed validation.

\subsection{Whirlwind v5.0}

No faults were detected when Whirlwind v4.0 was deployed so the latest revision focused on performance. This resulted in the introduction of custom schedulers and principals. Custom schedulers allowed the development of application or context specific schedulers. This introduced another mechanism to manage overload response. Principals came about as the categorization and prioritization code was being duplicated between message predicates and schedulers.

The byte code validator was removed in this version as it added significant overhead to developing Whirlwind--based services. It is unknown whether applying a more sophisticated validator as appeared in Kilim would have mitigated the cost.

\subsection{Reflections on the Evolution}

It is interesting that the evolution of Whirlwind drove it towards the actor model of concurrency. Initially the author was a strong advocate of event--driven architectures even in the context of multiple processors. Several attempts were made in an attempt to \emph{fix} this model in Whirlwind through techniques such as event coloring~\cite{Dabek02EventDriven} but with little practical success.

@@

\subsection{Influences}

Three major influences on Whirlwind are SEDA, Erlang and Kilim. Whirlwind incorporates; the load protection primitives from SEDA, fault management primitives and a supervisor library inspired by Erlang, and is an implementation of the actor model of concurrency for the Java programming language like Kilim. Whirlwind applications have a distinctly different flavor from any of these architectures due to different set of trade--offs.

Whirlwind eschews the shared memory model of SEDA and assumes processes are isolated and share no state. Unlike in Kilim and Erlang, ensuring the isolation of Whirlwind processes is partially the responsibility of the developer. A developer can inadvertently or deliberately share state between processes but in such a scenario, Whirlwind assumes that the developer knows what they are doing.

In SEDA, events about a single subject can be handled concurrently and can be reordered while in Whirlwind messages are processed sequentially in a FIFO order. Kilim and Erlang handle messages sequentially but Erlang need not receive messages from the inbox in a FIFO order.

Cohort scheduling, adaptive concurrency levels and message predicates are present in Whirlwind and SEDA but are not built in to Erlang or Kilim. None of these architectures make it easy to supply custom schedulers and no framework that we are aware of incorporates the principal abstraction.

Kilim and Erlang actors are independent threads of execution and can store state on the stack in between messages. Whirlwind and SEDA use a callback model and must explicitly store any continuation state in the process.

Lastly Erlang uses a specialized virtual machine to execute programs. The other architectures are all JVM based.

@@

There are three major changes planned for the next version of Whirlwind. The first change is to support Kilim~\cite{Srinivasan:08:Kilim} or Kilim--like processes executing on top of the Whirlwind runtime. The second is to investigate the possibility of dynamically organizing processes into groups at runtime. Lastly, Whirlwind should support alternative concurrency models for processes.

\subsection{Kilim integration}

Kilim has two main advantages over the current Whirlwind process implementation. Firstly, Kilim enforces the isolation constrains through a byte code validator. Earlier versions of Whirlwind attempted this with limited success.

Secondly, Kilim processes are lightweight threads that use automatic stack management~\cite{Adya02Cooperative} to maintain continuation data between handling messages. 
Whirlwind processes currently use complex state machines and manual stack management to achieve the same effect. Kilim attempts to optimize saving and restoring stack state when a process is paused or resumed. The degree of overhead is related to the number of frames on the stack when the process is paused or resumed. Experience with existing Whirlwind applications suggest that a paused process will only have a small number of active frames. The expectation is that the small performance degradation is an acceptable tradeoff for certain classes of process.

\subsection{Automatic grouping}

Simple heuristics combined with annotations and runtime instrumentation can automate the organization of processes into groups. Using automated grouping of processes will reduce the level of complexity for developing Whirlwind based applications. As groups are likely to be created on demand, processes will need to support migration from one group to another.

A typical heuristic would be to create a group when a large number of processes of a single type are instantiated. The processes are likely to have a high degree of code locality and with large numbers the system could take advantage of cohort scheduling making these processes an ideal group.

Annotations could be added to the processes to provide hints to the runtime system to guide grouping. The annotations would indicate the resource and concurrency requirements of the processes. Where annotations are not sufficient for the runtime to group processes appropriately, the developer can fall back on manual group configuration.

Automatic grouping would likely lead to the decoupling of the message predicate from the group. The message predicate is something that the developer will likely want to attach to specific sets of processes or even message flows between specific sets of processes.

\subsection{Alternative Concurrency Models}

At least one production application breaks the isolation assumptions of Whirlwind and shares state between processes. The shared state resulted from using an existing, high performance in--memory database library. The database library used fine--grain locking to mediate access to the shared state and maintains a high level of concurrency and throughput for the common case of no contention. Wrapping the database library in a single process resulted in a significant decrease in performance and was considered non--viable. The decision was made to share the state across several processes but move all the processes to a separate group and add error propagation paths between all processes.

A process that acquires a lock can influence the behavior of another process but that process is guaranteed to be in the same group. The groups scheduler and adaptive concurrency level controller were customized to take this into account when making resource allocation decisions.

Adding error propagation paths between all processes ensures that if any process fails, they all fail. This isolates the fault to the group but does mean that a single failure can result in the error propagating to a large number of dependent processes.

These steps ensured that the group is isolated from the rest of the system even if the processes within the group are not isolated from each other.

A pragmatic addition to Whirlwind would be to enable this design at the framework level. The framework would use one execution strategy when dealing with isolated message--driven processes and another when processes can also communicate via locks and shared memory. It is conceivable that this could be extended to processes that use shared transactional memory.

>>>>

\def\mytitle{Whirlwind: Overload protection, fault--tolerance and self--tuning in an Internet services platform}
\def\myauthor{Peter Donald, Samar Singh,  Somnath Ghosh}
\def\mykeywords{concurrency, software architecture, overload, fault-tolerance, self-tuning}

\documentclass[conference,a4paper,final]{IEEEtran}

\usepackage{cite}
\usepackage[pdftex]{graphicx,hyperref}
\usepackage{url}
\usepackage{epstopdf}
\usepackage[tight,footnotesize]{subfigure}

@@

@@

\author{\IEEEauthorblockN{\myauthor}
\IEEEauthorblockA{\textit{School of Computer Science and Electronic Engineering, La Trobe University,}\\
\textit{Melbourne, Australia}\\
\small \tt peter@realityforge.org, S.Singh@latrobe.edu.au, S.Ghosh@latrobe.edu.au}
}

\def\IEEEkeywordsname{Keywords}

@@

\begin{abstract}
Performance and availability are of critical importance when Internet services are integrated into emergency response management. Poor performance or service failure can result in severe economic, social or environmental cost. This paper presents Whirlwind, a software architecture that includes primitives for overload management and fault tolerance. A Whirlwind service is composed of a collection of isolated, independent, sequential processes that communicate through asynchronous message passing. If a process fails, the fault is contained within the process and a message is propagated to monitoring processes that may attempt to recover from the error. Processes are grouped with other processes that share similar resource, computation and concurrency requirements. Each group contains a scheduler and a thread pool that drives execution of processes within the group. The group may also define a message predicate that determines if a message posted to a process in the group is accepted. A rejected message typically signals overload and allows the application the chance to perform load shedding and avoid overcommitment of resources. Principals are shared between processes in different groups, enabling consistent prioritization and admission control across groups. The resource management policies are typically driven by feedback loops that monitor resource availability and system performance, and adjust tuning parameters to meet performance goals. Whirlwind evolved over a period of five fire seasons as part of emergency response software in Victoria, Australia.
\end{abstract}

\begin{keywords}
\mykeywords
\end{keywords}

@@

Whirlwind was born out of the need to create a reliable, scalable software architecture for the Incident Resource Information System (IRIS). IRIS tracks the dispatch and deployment of resources during an emergency response. The resources include everything from vehicles to personnel and the emergency situations have included wildfires, floods and virus outbreaks. IRIS is also responsible for generating the payments for personnel during emergency situations.

Internet traffic is bursty and popular Internet services are likely to be subject to periods of transient or sustained overload. This is intensified for Internet services used in emergency responses. Overload can be caused by unexpectedly high request rates or unexpected shifts in the characteristics of the request distribution. In IRIS the request rate increases dramatically during shift change over times or prior to payment cut--off dates. The characteristics of the requests also changed dramatically depending on whether the users were primarily interacting with the payment subsystem or the resource tracking subsystem.

@@

Whirlwind is based on the actor model of concurrency. It incorporates existing work on load management, fault--tolerance, scheduling and self--tuning and synthesizes these elements into a consistent framework. Whirlwind was developed for the Java Virtual Machine (JVM), primarily targeting the Java programming language. The architecture needed to take advantage of the increased number of processing units in modern hardware.

Each iteration of Whirlwind has undergone an extensive performance evaluation before being rolled out into production. Space does not permit presenting a a detailed performance report. Suffice to say that Whirlwind, at least in certain configurations, met the original performance goals and compares favorably with other popular production frameworks.

@@

Past research has tended to focus on one categorization axis rather than categorizing a request along multiple dimensions. It is probable that this is done as it makes the mapping between a categorization and a priority simple; \emph{Premium} users are always prioritized above \emph{Basic} users, \emph{Buy} requests are always prioritized above \emph{Browse} requests, while during overload light--weight requests are always prioritized above heavy--weight requests etc.

@@

A fault tolerant system is one able to recover from errors and restore normal operation. If there is a decrease in service quality, the degradation is proportional to the severity of the fault. Fault tolerance is achieved by expecting faults and designing the system so that it behaves reasonably in the presence of faults.

@@

@@

Erlang code is executed by a specialized runtime. However other actor implementations (i.e. SALSA~\cite{Varela:SALSA}, Scala's Actor library~\cite{Haller:07:ScalaActors}, Kilim~\cite{Srinivasan:08:Kilim}) target a general purpose virtual machine such as the JVM. These implementations use the type system or code modification techniques to verify that the actors conform to certain rules (i.e. no access to shared memory) and to integrate the actors with a runtime library. The runtime library provides execution environment for the actors and maps the actors to threads using the 1--to--1 model~\cite{Varela:SALSA}, M--to--N model~\cite{Srinivasan:08:Kilim,armstrong96erlang} or a combination of both approaches~\cite{Haller:07:ScalaActors}.

@@

@@

Processes are isolated and do not share memory. To communicate with a process a message is posted to the process. The ownership of the message contents is transferred to the receiving process. Message delivery is asynchronous and messages are placed in a process--local queue or inbox. Message transmission is asynchronous to ensure that the sender is isolated from the receiver and the sender can not be blocked, or delayed by a slow receiver. It should be noted that message passing in Whirlwind obeys the \emph{happened--before} ordering of events.

@@

@@

Most Internet services support many concurrent requests and many of these requests perform similar computations. Identifying requests that are at a similar stage of processing and executing a ``cohort'' of them sequentially increases code and data locality. This in turn increases performance by reducing cache misses~\cite{Larus:2002:Cohort,welsh03Adaptive}. Groups are designed to identify processes that have a high degree of code and data locality and thus the group defines an ideal ``cohort''.

@@

The post operation will raise an exception when a message is rejected thus allowing the source process the opportunity to attempt alternative action. A HTTP service may first attempt to deliver a media rich page. If this message is rejected, the service attempts to deliver a lower resolution page as presented in Figure~\ref{figure:http-server}(b). This allows the service to deliver progressively lower resolution content as the load on the service increases thus avoiding overload~\cite{abdelzaher99web}. SEDA demonstrated the effectiveness of the message predicate primitive in overload management~\cite{welsh03Adaptive}.

@@

Whirlwind provides several controller implementations that dynamically adjust the concurrency level of a group. Periodically the controller observes performance characteristics and adjusts the pool size in an attempt to meet throughput and response time goals. The developer can also specify a custom controller or use one of the predefined controllers.

@@

As in Erlang, the supervisor processes are typically organized into a tree such that all but the root supervisor process is monitored by another supervisor. If a supervisor fails to recover from a child failure then it can in turn fail and force the parent supervisor process to attempt error recovery. For the system to fail, all supervisors in the tree must fail to recover from the error.

@@

The best way to get a feel for the architecture is to look at the design for a Whirlwind application. This section describes the design of the HTTP server component used in IRIS. Figure~\ref{figure:http-server}(a) presents a basic design of subsystem responsible for processing HTTP connections.

\begin{figure*}[htbp]
\centering
\includegraphics[width=\linewidth]{http-server.eps}
\caption{(a) The subsystem responsible for processing HTTP connections. All of the groups contain processes of a single type except for the ``Resource Handler'' group that includes processes of type ``List'' and ``Show'' for listing a set of resources and presenting detailed information about a single resource respectively. For the sake of simplicity, the message paths are presented as flowing in one direction but all paths are bidirectional except that the handlers that communicate directly with \emph{HttpConnection} processes. (b) The message predicate attached to the \emph{FileHandler} group rejects a message requesting the high quality version of a file but accepts a message requesting a less resource intensive, low quality version of a file.}
\label{figure:http-server}
\end{figure*}

Each connection to the HTTP server resulted in the creation of three processes; a \verb+SocketConnection+ process to manage the underlying socket, a \verb+HttpConnection+ process to decode and encode HTTP messages and manage the HTTP connection state, and a \verb+Dispatcher+ that routes the HTTP requests to specific handlers based on the contents of the headers, method and uri of the request. Each request submitted to the server results in the \verb+Dispatcher+ creating a \emph{handler} process. A handler is responsible for delivering content from the filesystem or dynamically generating content based based on values retrieved from a database.

The handler will usually generate one or more processes while handling the request. For example, the \verb+FileHandler+ process creates an \verb+AsyncFile+ process while the \verb+ListResources+ process creates a \verb+DbConnection+ process. Handlers share a group if they utilize similar resources and load shedding treats these handlers similarly. However, most handlers and other other processes in the system will have a single group per process type.

Most Whirlwind applications will only place messages predicates on a few specific groups. The HTTP server defines a message predicate on the \verb+Dispatcher+ group and on each of the handler groups. If a message predicate attached to a handler group rejects a message then the dispatcher may attempt to send a less resource intensive request to the handler, as presented in Figure~\ref{figure:http-server}(b). If the low resolution message is rejected then the \verb+Dispatcher+ returns a ``503 Service Unavailable'' response to the client and closes the connection.

@@

The creation of a new handler for every request rather than handler reuse is deliberate. This design results in all the processes in Figure~\ref{figure:http-server} being  associated with a single principal thus allowing consistent treatment by schedulers and message predicates across the whole system.

The prioritization of principals takes into account whether the session is for an authenticated user and if the user is in the \emph{admin} role. Administrators are given the highest priority while unauthenticated users are given the lowest priority.

@@

Whirlwind has gone through 5 major revisions. Initially Whirlwind was a shared--memory, event--driven architecture heavily influenced by SEDA~\cite{welsh03Adaptive}. At the end of each fire season, the architecture of Whirlwind was re--evaluated and changes made to avoid problems experienced in the production environment.

By Whirlwind v4.0 the architecture had evolved to use the actor model of concurrency and had incorporated the Erlang model of fault propagation and recovery. The concurrency and isolation constraints of processes were enforced by a byte code validator not dissimilar to the byte code weaver in Kilim~\cite{Srinivasan:08:Kilim}.

@@

@@

There are two major changes planned for the next version of Whirlwind. The first change is to support Kilim~\cite{Srinivasan:08:Kilim} or Kilim--like processes executing on top of the Whirlwind runtime. The second is to investigate the possibility of dynamically organizing processes into groups at runtime.

Kilim processes are lightweight threads that use automatic stack management~\cite{Adya02Cooperative} to maintain continuation data between handling messages compared to more complex manual stack management present in Whirlwind processes. Kilim attempts to optimize saving and restoring stack state when a process is paused or resumed. The degree of overhead is related to the number of frames on the stack when the process is paused or resumed. Experience with existing Whirlwind applications suggest that a paused process will only have a small number of active frames. The expectation is that the small performance degradation is an acceptable tradeoff for certain classes of process.

Automatic grouping of processes will reduce the level of complexity for developing Whirlwind based applications. As groups are likely to be created on demand, processes will need to support migration from one group to another. Automatic grouping would also lead to the decoupling of the message predicate from the group. Simple heuristics combined with annotations and runtime instrumentation can automate the organization of processes into groups with manual grouping possible when heuristics perform poorly.
